---
output:
  pdf_document: default
  html_document: default
---
# Cortical Thickness Analysis

This section runs the stats for average (by hemisphere) Cortical Thickness calculated with Freesurfer

```{r set_up, message=FALSE, warning=FALSE}

#load libraries
library(tidyverse)
library(broom)
library(lmerTest)
library(tableone)


#bring in data 
df <- read_csv('../generated_csvs/STOPPD_participantsCT_20181111.csv') #generated by 05_STOPPD_error in prepartion of Jason Lerch meeting


```


```{r}
#make sure that STUDYID is an interger not a number
  df$STUDYID <- as.character(df$STUDYID)

#make sure that dateDiff is a number, not an interger
  df$dateDiff <- as.numeric(df$dateDiff)

# label the randomization variable  
df$RandomArm <- factor(df$randomization, 
                       levels = c("O", "P"),
                       labels = c("Olanzapine", "Placebo"))

RandomArmColors = c("#e6ab02", "#386cb0")

# set category levels so that RCT and Relapse are at the top
df <- df %>%
  mutate(category = factor(category, levels = c("RCT","Relapse", "Off protocol")))

#restructure data for RCT completers' only (N=40)
  RCT_CT <- df %>%
    filter(category == "RCT") 



#write out clean dataframe
 # write.csv(RCT_CT, '../generated_data/df_leftCT.csv', row.names=FALSE)

```



#### baseline measures (table1 part 1)

```{r}
CreateTableOne(data = df,
               strata = "randomization",
               vars = c("category", "LThickness_01", "RThickness_01", "LSurfArea_01", "RSurfArea_01"))
```

```{r}
CreateTableOne(data = df,
               vars = c("category", "LThickness_01", "RThickness_01", "LSurfArea_01", "RSurfArea_01"))
```

#### baseline stats (part 2)

```{r}

df %>%
  select(randomization, LThickness_01, RThickness_01, LSurfArea_01, RSurfArea_01) %>%
  gather(thick, mm, -randomization) %>%
  group_by(thick) %>%
  do(tidy(t.test(mm~randomization, data = .))) %>%
  knitr::kable(caption = "t.test for baseline group differences")

```

## RCT only

```{r RCT_LCT_boxplot}
RCT_CT %>% count(randomization) %>% knitr::kable() 

```


```{r RCT_LCT_stats}

fit_all <- lmer(LThickness_change ~ RandomArm + sex + age + (1|site), data= RCT_CT)
summary(fit_all) 

fit_all <- lmer(LThickness_change ~ RandomArm + sex + age + (1|site), data= filter(RCT_CT, age > 50))
summary(fit_all) 
  
```
```{r RCT_RCT_stats}

fit_all <- lmer(RThickness_change ~ RandomArm + sex + age + (1|site), data= RCT_CT)
summary(fit_all)  

fit_all <- lmer(RThickness_change ~ RandomArm + sex + age + (1|site), data= filter(RCT_CT, age > 50))
summary(fit_all)  
```

### looking at the same thing for Right CT


```{r RCT_CT_facet_boxplot_fig2A, fig.height=3.5, fig.width = 5}
#boxplot of difference in thickness (y axis) by randoMR_exclusion == "No"mization group (x axis)
RCT_CT %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere"))) %>%
ggplot(aes(x= RandomArm, y = mm, fill = RandomArm)) + 
     geom_boxplot(outlier.shape = NA, alpha = 0.0001) + 
     geom_dotplot(binaxis = 'y', stackdir = 'center', binwidth = 0.005) +
     geom_hline(yintercept = 0) +
     labs(x = NULL, y = "Change in Cortical Thickness (mm)") +
     scale_fill_manual(values = RandomArmColors) +
     scale_shape_manual(values = c(21)) +
     facet_wrap(~ ThickChange) +
     theme_bw() +
     theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) 

```

```{r RCT_CT_facet_boxplot_fig2Aa, fig.height=3.5, fig.width = 5}
#boxplot of difference in thickness (y axis) by randoMR_exclusion == "No"mization group (x axis)
RCT_CT %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere"))) %>%
ggplot(aes(x= RandomArm, y = mm, fill = RandomArm)) + 
     geom_boxplot(outlier.shape = NA, alpha = 0.0001) + 
     geom_dotplot(binaxis = 'y', stackdir = 'center', binwidth = 0.005) +
     geom_hline(yintercept = 0) +
     labs(x = NULL, y = "Change in Cortical Thickness (mm)") +
     scale_fill_manual(values = RandomArmColors) +
     scale_shape_manual(values = c(21)) +
     facet_wrap(~ ThickChange) +
     theme_bw() +
     theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) 
```
```{r}
RCT_CT %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere"))) %>%
ggplot(aes(x= RandomArm, y = mm, fill = factor(age > 50))) + 
     geom_boxplot(outlier.shape = NA, alpha = 0.0001) + 
     geom_dotplot(binaxis = 'y', stackdir = 'center', binwidth = 0.005) +
     geom_hline(yintercept = 0) +
     labs(x = NULL, y = "Change in Cortical Thickness (mm)") +
     scale_fill_manual(values = RandomArmColors) +
     scale_shape_manual(values = c(21)) +
     facet_wrap(~ ThickChange) +
     theme_bw() +
     theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) 

```


## RCT & Relapse (with time as factor)

```{r the-ns}
#restructure data for RCT & Relapse participants (N=72)
  RCTRelapse_LCT <- df %>%
    gather(thick_oldcolname, thickness, LThickness_01, LThickness_02) %>%
    mutate(model_days = if_else(thick_oldcolname == "LThickness_01", 1, dateDiff)) %>%
    mutate(category = factor(category, levels = c("RCT","Relapse", "Off protocol")),
           hemi = "Left Hemisphere")

RCTRelapse_LCT %>% filter(model_days == 1) %>% count(RandomArm, offLabel) %>% knitr::kable() 

RCTRelapse_LCT_sensitivety <- RCTRelapse_LCT %>% 
  filter(category != "Off protocol" )  

RCTRelapse_LCT_sensitivety %>% filter(model_days == 1) %>% count(RandomArm, offLabel) %>% knitr::kable() 
```


```{r RCTRelapse_LCT_plot_fig2CL, warning = FALSE, fig.height=4, fig.width = 4}
RCTRelapse_LCT %>%
  ggplot(aes(x=model_days, y=thickness, fill = RandomArm)) + 
  geom_point(aes(shape = category)) + 
  geom_line(aes(group=STUDYID, color = RandomArm), alpha = 0.5) + 
  geom_smooth(aes(color = RandomArm), method="lm", fill = "grey40") +
  labs(x = "Days between MRIs", y = "Cortical Thickness (mm)", colour = NULL) +
  scale_color_manual(values = RandomArmColors) +
  scale_fill_manual(values = RandomArmColors) +
  scale_shape_manual(values = c(21:23)) +
  scale_y_continuous(limits = c(2.1,2.65)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
  facet_wrap(~hemi)
```




```{r RCTRelapse_LCT_statswsite, warning = FALSE}  
#run mixed linear model, with covariates
  fit_all <- lmer(thickness ~ RandomArm*model_days + age + sex + (1|site) + (1|STUDYID), data= RCTRelapse_LCT)
  summary(fit_all)
  
  fit_all <- lmer(thickness ~ RandomArm*model_days + age + sex + (1|site) + (1|STUDYID), data= filter(RCTRelapse_LCT, age > 50))
  summary(fit_all)

```



```{r RCTRelapse_LCT_statswsite_sense, warning = FALSE}  

#run mixed linear model, with covariates
  fit_all <- lmer(thickness ~ RandomArm*model_days + age + sex + (1|site) + (1|STUDYID), data= RCTRelapse_LCT_sensitivety)
  summary(fit_all)
  
  fit_all <- lmer(thickness ~ RandomArm*poly(model_days,2) + age + sex + (1|site) + (1|STUDYID), data= RCTRelapse_LCT_sensitivety)
  summary(fit_all)

    fit_all <- lmer(thickness ~ RandomArm*model_days + age + sex + (1|site) + (1|STUDYID), data= filter(RCTRelapse_LCT_sensitivety, age > 50))
  summary(fit_all)
```

### Running the right hemisphere RCTRelapse

```{r RCTRelapse_RCT_plot, warning = FALSE}

#restructure data for RCT & Relapse participants (N=72)
  RCTRelapse_RCT <- df %>%
    gather(thick_oldcolname, thickness, RThickness_01, RThickness_02) %>%
    mutate(model_days = if_else(thick_oldcolname == "RThickness_01", 1, dateDiff)) %>%
    mutate(category = factor(category, levels = c("RCT","Relapse", "Off protocol")),
           hemi = "Right Hemisphere")

RCTRelapse_RCT_sensitivety <- RCTRelapse_RCT %>% 
  filter(category != "Off protocol" )  

```

```{r RCTRelapse_RCT_plot_fig2CR, warning = FALSE, fig.height=4, fig.width = 3.75}
RCTRelapse_RCT %>%
  ggplot(aes(x=model_days, y=thickness, fill = RandomArm)) + 
  geom_point(aes(shape = category)) + 
  geom_line(aes(group=STUDYID, color = RandomArm), alpha = 0.5) + 
  geom_smooth(aes(color = RandomArm), method="lm", fill = "grey40") +
  labs(x = "Days between MRIs", y = "", colour = NULL) +
  scale_color_manual(values = RandomArmColors) +
  scale_fill_manual(values = RandomArmColors) +
  scale_shape_manual(values = c(21:23)) +
  scale_y_continuous(limits = c(2.1,2.65), breaks = NULL) +
  theme_bw()  +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
  facet_wrap(~hemi)
```
```{r RCTRelapse_RCT_statswsite2, warning = FALSE}  
#run mixed linear model, with covariates
  fit_all <- lmer(thickness ~ RandomArm*model_days + sex + age + (1|site) + (1|STUDYID), data= RCTRelapse_RCT)
  summary(fit_all) 
  
  fit_all <- lmer(thickness ~ RandomArm*poly(model_days,2) + sex + age + (1|site) + (1|STUDYID), data= RCTRelapse_RCT)
  summary(fit_all)  

    fit_all <- lmer(thickness ~ RandomArm*model_days + sex + age + (1|site) + (1|STUDYID), data= filter(RCTRelapse_RCT, age > 50))
  summary(fit_all) 

```


```{r RCTRelapse_RCT_statswsite2_sense, warning = FALSE}  
#run mixed linear model, with covariates
  fit_all <- lmer(thickness ~ RandomArm*model_days + sex + age + (1|site) + (1|STUDYID), data= RCTRelapse_RCT_sensitivety)
  summary(fit_all)  


```

### Playing with other ways to layout the plots


```{r RCTRelapse_LCT_plot_2part1, warning = FALSE}
RCTRelapse_LCT %>%
  ggplot(aes(x=model_days, y=thickness, colour = category)) +
  geom_point() +
  geom_line(aes(group=STUDYID), alpha = 0.5) +
  ggtitle("Cortical thickness in left hemisphere over time") +
  labs(x = "Days between MRIs", y = "Cortical Thickness (mm)", colour = NULL) +
  theme_minimal() + facet_wrap(~ RandomArm) +
  scale_colour_discrete(direction = -1)
```

```{r RCTRelapse_RCT_plot_2part2, warning = FALSE}
RCTRelapse_LCT %>%
  mutate(Outcome = case_when(category == "Off protocol" ~ "non-completer",
                             category == "Relapse" ~ "non-completer",
                             category == "RCT"~ "completer")) %>%
  ggplot(aes(x=model_days, y=thickness, fill = category)) +
  geom_point() +
  geom_line(aes(group=STUDYID), alpha = 0.5) +
  ggtitle("Cortical thickness in left hemisphere over time") +
  labs(x = "Days between MRIs", y = "Cortical Thickness (mm)", colour = NULL) +
  theme_bw() + facet_wrap(~ RandomArm) +
  scale_colour_manual(values = c("black", "white", "grey"))
```

## Post-Hoc - looking at subgroups

plotting change for all participants


```{r}
df %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere")),
         Outcome = case_when(category == "Off protocol" ~ "non-completer",
                             category == "Relapse" ~ "non-completer",
                             category == "RCT"~ "completer")) %>%
  ggplot(aes(x= RandomArm, y = mm, fill = Outcome)) + 
     geom_boxplot(outlier.shape = NA) + 
     geom_dotplot(binaxis = 'y', stackdir = 'center', 
                  position=position_dodge(0.8), binwidth = 0.005) +
     geom_hline(yintercept = 0) +
     xlab(NULL) +
     ylab("Change in Cortical Thickness (mm)") +
     theme_bw() +
     scale_fill_manual(values = c('white','grey')) +
     facet_wrap(~ThickChange)
```

```{r}
df %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere")),
         Outcome = case_when(category == "Off protocol" ~ "non-completer",
                             category == "Relapse" ~ "non-completer",
                             category == "RCT"~ "completer")) %>%
  filter(category != "Off protocol") %>%
  group_by(ThickChange, RandomArm, category) %>%
  do(tidy(t.test(.$mm, mu = 0, alternative = "two.sided"))) %>%
  knitr::kable(digits = 3)
```
## Exploratory within Treatment Arm tests

```{r}
df %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere")),
         Outcome = case_when(category == "Off protocol" ~ "non-completer",
                             category == "Relapse" ~ "non-completer",
                             category == "RCT"~ "completer")) %>%
  filter(category != "Off protocol") %>%
  group_by(ThickChange, RandomArm) %>%
  do(tidy(t.test(mm~category, var.equal = FALSE, data = .))) %>%
  knitr::kable(digits = 3)
```
```{r}
df %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere")),
         Outcome = case_when(category == "Off protocol" ~ "non-completer",
                             category == "Relapse" ~ "non-completer",
                             category == "RCT"~ "completer")) %>%
  filter(category != "Off protocol") %>%
  group_by(ThickChange, RandomArm) %>%
  do(tidy(lm(mm~category + age + sex + site, var.equal = FALSE, data = .))) %>%
  filter(term == "categoryRelapse") %>%
  knitr::kable(digits = 3)
```
```{r}
df %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere")),
         MRdays = as.numeric(dateDiff)) %>%
  filter(category != "Off protocol") %>%
  group_by(ThickChange, RandomArm) %>%
  do(tidy(lm(mm~MRdays, data = .))) %>%
  filter(term == "MRdays") %>%
  knitr::kable(digits = 3)
```
```{r fig.height=6, fig.width=6}
df %>%
  gather(TCT, mm, LThickness_change, RThickness_change) %>%
  mutate(ThickChange = factor(TCT, levels = c("LThickness_change", "RThickness_change"),
                              labels = c("Left Hemisphere", "Right Hemisphere")),
         MRdays = as.numeric(dateDiff)) %>%
  filter(category != "Off protocol") %>%
  ggplot(aes(x = MRdays, y = mm)) +
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_grid(RandomArm ~ ThickChange)
```

## Exporatory ROI Analysis..

Running the RCT analysis ROI-wise with FDR correction. 

```{r }
library(ggseg)
library(magrittr)
library(dplyr)
library(broom)
```




```{r }
RCT_ROIwise <- RCT_CT %>%
  gather(elabel, change_mm, ends_with('_thickavg_change')) %>%
  group_by(elabel) %>%
  do(tidy(lm(change_mm ~ RandomArm + sex + age + site, data= .))) %>%
  ungroup() %>% group_by(term) %>%
  mutate(p_FDR = p.adjust(p.value, method = 'fdr'))

RCT_ROIwise_supptable = RCT_ROIwise %>%
  filter(p_FDR < 0.06) %>%
  arrange(p.value) %>%
  mutate(ROI = str_replace(elabel, '_thickavg_change', '')) %>%
  ungroup() %>%
  select(ROI, estimate, std.error, statistic, p_FDR)

RCT_ROIwise_supptable %>% write_csv('../generated_csvs/supptable4_thickbyROI.csv')

RCT_ROIwise_supptable %>%
  knitr::kable(digits = 3)
```
```{r roiwise_cort_thickness_plot_supplfig}
library(viridis)
RCT_ROIwise %>%
  ungroup() %>%
  filter(term == "RandomArmPlacebo") %>%
  mutate(label = str_replace(elabel, '_thickavg_change', '') %>%
           str_replace('L','lh') %>%
           str_replace('R','rh'),
         is_sig = if_else(p_FDR < 0.055, "sig", NA_character_)) %>%
  inner_join(atlas.info$data[2][[1]], by = "label") %>%
  ggseg(atlas="dkt", mapping=aes(fill=statistic, color = is_sig)) +
  scale_fill_viridis() +
  scale_color_manual(values = c("black", NULL)) + 
  labs(color = NULL, fill = "t-statistic")
  
```

### Figure Caption (brain plots) 

Mapping the effect of olanzapine vs placebo on cortical thinning over 36 weeks in participant who remained clinically stable. The color scale represents the t-statistic for the effect of treatment (Placebo vs Olanzapine) where brighter colors represents greater cortical thinning. Areas outlined in black are those where the treatment effects was significant after correction for multiple comparisons (across 68 brain regions) using False Discovery Rate. 

